---
title: "Evaluating Exercise by Machine Learning"
author: "Lakshmanan Arasakumar"
output: pdf_document
---

###Overview

Machine Learning is the process of using algorithms to learn from data. Perhaps the most important aspect of any machine learning problem is the rather human process of determing what we are trying to learn about.

In the following study we will discuss the excercise data. They investigated the use of computing to evaluate "proper" exercise form (possibly allowing computers to replace personal trainers to help us become better, faster, stronger.

###Data Loading and Processing

The first step is to import the data and to verify that the training data and the test data are identical.

```{r }

library(AppliedPredictiveModeling)
library(caret)
library(rattle)
library(rpart.plot)
library(randomForest)
library(RCurl)

# Download data.

data <- getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", ssl.verifypeer=0L, followlocation=1L)
writeLines(data,'temp.csv')
df_training<-read.csv('temp.csv', na.strings=c("NA",""), header=TRUE)
colnames_train <- colnames(df_training)

data1 <- getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", ssl.verifypeer=0L, followlocation=1L)
writeLines(data1,'temp1.csv')
df_testing<-read.csv('temp1.csv', na.strings=c("NA",""), header=TRUE)
colnames_test <- colnames(df_testing)

# Verify that the column names (excluding classe and problem_id) are identical in the training and test set.
all.equal(colnames_train[1:length(colnames_train)-1], colnames_test[1:length(colnames_train)-1])

```

Having verified that the schema of both the training and testing sets are identical (excluding the final column representing the A-E class), I decided to eliminate both NA columns and other extraneous columns.

````{r}
# Count the number of non-NAs in each col.
nonNAs <- function(x) {
    as.vector(apply(x, 2, function(x) length(which(!is.na(x)))))
}

# Build vector of missing data or NA columns to drop.
colcnts <- nonNAs(df_training)
drops <- c()
for (cnt in 1:length(colcnts)) {
    if (colcnts[cnt] < nrow(df_training)) {
        drops <- c(drops, colnames_train[cnt])
    }
}

# Drop NA data and the first 7 columns as they're unnecessary for predicting.
df_training <- df_training[,!(names(df_training) %in% drops)]
df_training <- df_training[,8:length(colnames(df_training))]

df_testing <- df_testing[,!(names(df_testing) %in% drops)]
df_testing <- df_testing[,8:length(colnames(df_testing))]

# Show remaining columns.
colnames(df_training)
colnames(df_testing)

````

Given that we're already supplied with the raw sensor data, there's no need for Level 1 processing. However, while being careful not to overfit, some Level 2 processing is certainly worth attempting. First, check for covariates that have virtually no variablility.


```{r}
nsv <- nearZeroVar(df_training, saveMetrics=TRUE)
nsv
```

Given that all of the near zero variance variables (nsv) are FALSE, there's no need to eliminate any covariates due to lack of variablility.

###ALGORITHM

We were provided with a large training set (19,622 entries) and a small testing set (20 entries). Instead of performing the algorithm on the entire training set, as it would be time consuming and wouldn't allow for an attempt on a testing set, I chose to divide the given training set into four roughly equal sets, each of which was then split into a training set (comprising 60% of the entries) and a testing set (comprising 40% of the entries).

```{r}
# Divide the given training set into 4 roughly equal sets.
set.seed(666)
ids_small <- createDataPartition(y=df_training$classe, p=0.25, list=FALSE)
df_small1 <- df_training[ids_small,]
df_remainder <- df_training[-ids_small,]
set.seed(666)
ids_small <- createDataPartition(y=df_remainder$classe, p=0.33, list=FALSE)
df_small2 <- df_remainder[ids_small,]
df_remainder <- df_remainder[-ids_small,]
set.seed(666)
ids_small <- createDataPartition(y=df_remainder$classe, p=0.5, list=FALSE)
df_small3 <- df_remainder[ids_small,]
df_small4 <- df_remainder[-ids_small,]
# Divide each of these 4 sets into training (60%) and test (40%) sets.
set.seed(666)
inTrain <- createDataPartition(y=df_small1$classe, p=0.6, list=FALSE)
df_small_training1 <- df_small1[inTrain,]
df_small_testing1 <- df_small1[-inTrain,]
set.seed(666)
inTrain <- createDataPartition(y=df_small2$classe, p=0.6, list=FALSE)
df_small_training2 <- df_small2[inTrain,]
df_small_testing2 <- df_small2[-inTrain,]
set.seed(666)
inTrain <- createDataPartition(y=df_small3$classe, p=0.6, list=FALSE)
df_small_training3 <- df_small3[inTrain,]
df_small_testing3 <- df_small3[-inTrain,]
set.seed(666)
inTrain <- createDataPartition(y=df_small4$classe, p=0.6, list=FALSE)
df_small_training4 <- df_small4[inTrain,]
df_small_testing4 <- df_small4[-inTrain,]

```
Based on both the process outlined in Section 5.2 of the aforementioned paper and the concensus in the coursera discussion forums, I chose two different algorithms via the caret package: classification trees (method = rpart) and random forests (method = rf).

###PARAMETERS
We decided to try classification trees "out of the box" and then introduce preprocessing and cross validation.

While we also considered applying "out of the box" random forest models, some of the horror stories contributed to the coursera discussion forums regarding the lengthy processing times for random forest models convinced me to only attempt random forests with cross validation and, possibly, preprocessing.

###EVALUATION
####Classification Tree

First, the "out of the box" classification tree:

```{r}
# Train on training set 1 of 4 with no extra features.
set.seed(666)
modFit <- train(df_small_training1$classe ~ ., data = df_small_training1, method="rpart")
print(modFit, digits=3)
print(modFit$finalModel, digits=3)
fancyRpartPlot(modFit$finalModel)
```

```{r}
# Run against testing set 1 of 4 with no extra features.
predictions <- predict(modFit, newdata=df_small_testing1)
print(confusionMatrix(predictions, df_small_testing1$classe), digits=4)
# Train on training set 1 of 4 with only preprocessing.
set.seed(666)
modFit <- train(df_small_training1$classe ~ .,  preProcess=c("center", "scale"), data = df_small_training1, method="rpart")
print(modFit, digits=3)
# Train on training set 1 of 4 with only cross validation.
set.seed(666)
modFit <- train(df_small_training1$classe ~ .,  trControl=trainControl(method = "cv", number = 4), data = df_small_training1, method="rpart")
print(modFit, digits=3)

# Train on training set 1 of 4 with both preprocessing and cross validation.
set.seed(666)
modFit <- train(df_small_training1$classe ~ .,  preProcess=c("center", "scale"), trControl=trainControl(method = "cv", number = 4), data = df_small_training1, method="rpart")
print(modFit, digits=3)

# Run against testing set 1 of 4 with both preprocessing and cross validation.
predictions <- predict(modFit, newdata=df_small_testing1)
print(confusionMatrix(predictions, df_small_testing1$classe), digits=4)

```
The impact of incorporating both preprocessing and cross validation appeared to show some minimal improvement (accuracy rate rose from 0.531 to 0.552 against training sets). However, when run against the corresponding testing set, the accuracy rate was identical (0.5584) for both the "out of the box" and the preprocessing/cross validation methods.

###Random Forest
Decided to assess the impact/value of including preprocessing.

```{r}
# Train on training set 1 of 4 with only cross validation.
set.seed(666)
modFit <- train(df_small_training1$classe ~ ., method="rf", trControl=trainControl(method = "cv", number = 4), data=df_small_training1)
print(modFit, digits=3)
# Run against testing set 1 of 4.
predictions <- predict(modFit, newdata=df_small_testing1)
print(confusionMatrix(predictions, df_small_testing1$classe), digits=4)
# Run against 20 testing set provided by Professor Leek.
print(predict(modFit, newdata=df_testing))
# Train on training set 1 of 4 with only both preprocessing and cross validation.
set.seed(666)
modFit <- train(df_small_training1$classe ~ ., method="rf", preProcess=c("center", "scale"), trControl=trainControl(method = "cv", number = 4), data=df_small_training1)
print(modFit, digits=3)
# Run against testing set 1 of 4.
predictions <- predict(modFit, newdata=df_small_testing1)
print(confusionMatrix(predictions, df_small_testing1$classe), digits=4)
# Run against 20 testing set provided by Professor Leek.
print(predict(modFit, newdata=df_testing))
```

Preprocessing actually lowered the accuracy rate from 0.955 to 0.954 against the training set. However, when run against the corresponding set, the accuracy rate rose from 0.9689 to 0.9714 with the addition of preprocessing. Thus I decided to apply both preprocessing and cross validation to the remaining 3 data sets.

```{r}
# Train on training set 2 of 4 with only cross validation.
set.seed(666)
modFit <- train(df_small_training2$classe ~ ., method="rf", preProcess=c("center", "scale"), trControl=trainControl(method = "cv", number = 4), data=df_small_training2)
print(modFit, digits=3)
# Run against testing set 2 of 4.
predictions <- predict(modFit, newdata=df_small_testing2)
print(confusionMatrix(predictions, df_small_testing2$classe), digits=4)
# Run against 20 testing set provided by Professor Leek.
print(predict(modFit, newdata=df_testing))
# Train on training set 3 of 4 with only cross validation.
set.seed(666)
modFit <- train(df_small_training3$classe ~ ., method="rf", preProcess=c("center", "scale"), trControl=trainControl(method = "cv", number = 4), data=df_small_training3)
print(modFit, digits=3)
# Run against testing set 3 of 4.
predictions <- predict(modFit, newdata=df_small_testing3)
print(confusionMatrix(predictions, df_small_testing3$classe), digits=4)
# Run against 20 testing set provided by Professor Leek.
print(predict(modFit, newdata=df_testing))
# Train on training set 4 of 4 with only cross validation.
set.seed(666)
modFit <- train(df_small_training4$classe ~ ., method="rf", preProcess=c("center", "scale"), trControl=trainControl(method = "cv", number = 4), data=df_small_training4)
print(modFit, digits=3)
# Run against testing set 4 of 4.
predictions <- predict(modFit, newdata=df_small_testing4)
print(confusionMatrix(predictions, df_small_testing4$classe), digits=4)

# Run against 20 testing set provided by Professor Leek.
print(predict(modFit, newdata=df_testing))

```

####Out of Sample Error
Random Forest (preprocessing and cross validation) Testing Set 1: 1 - .9714 = 0.0286
Random Forest (preprocessing and cross validation) Testing Set 2: 1 - .9634 = 0.0366
Random Forest (preprocessing and cross validation) Testing Set 3: 1 - .9655 = 0.0345
Random Forest (preprocessing and cross validation) Testing Set 4: 1 - .9563 = 0.0437
Since each testing set is roughly of equal size, I decided to average the out of sample error rates derived by applying the random forest method with both preprocessing and cross validation against test sets 1-4 yielding a predicted out of sample rate of 0.03585.


###Conclusion

We received three separate predictions by appling the 4 models against the actual 20 item training set:

A) Accuracy Rate 0.0286 Predictions: B A A A A E D B A A B C B A E E A B B B

B) Accuracy Rates 0.0366 and 0.0345 Predictions: B A B A A E D B A A B C B A E E A B B B

C) Accuracy Rate 0.0437 Predictions: B A B A A E D D A A B C B A E E A B B B

Since Professor Leek is allowing 2 submissions for each problem, I decided to attempt with the two most likely prediction sets: option A and option B.

Since options A and B above only differed for item 3 (A for option A, B for option B), I subimitted one value for problems 1-2 and 4-20, while I submitted two values for problem 3. For problem 3, I was expecting the automated grader to tell me which answer (A or B) was correct, but instead the grader simply told me I had a correct answer. All other answers were also correct, resulting in a score of 100%.




